---
title: "PCA"
author: "Sophia Leeman"
date: "2023-02-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(factoextra)
library(tidyverse)
library(corrplot)
```

This document contains preliminary PCA clustering of the Welltory Health Dataset

Import csvs

```{r}
bp <- read.csv("/Users/sophialeeman/Documents/BME 580/hrv-covid19/data/blood_pressure.csv")
hr <- read.csv("/Users/sophialeeman/Documents/BME 580/hrv-covid19/data/heart_rate.csv")
hrv <- read.csv("/Users/sophialeeman/Documents/BME 580/hrv-covid19/data/hrv_measurements.csv")
p <- read.csv("/Users/sophialeeman/Documents/BME 580/hrv-covid19/data/participants.csv")
scale <- read.csv("/Users/sophialeeman/Documents/BME 580/hrv-covid19/data/scales_description.csv")
sleep <- read.csv("/Users/sophialeeman/Documents/BME 580/hrv-covid19/data/sleep.csv")
survey <- read.csv("/Users/sophialeeman/Documents/BME 580/hrv-covid19/data/surveys.csv")
wear <- read.csv("/Users/sophialeeman/Documents/BME 580/hrv-covid19/data/wearables.csv")
weather <- read.csv("/Users/sophialeeman/Documents/BME 580/hrv-covid19/data/weather.csv")
```

View csvs
```{r}
head(bp, 2)
summary(bp)
head(hr, 2)
summary(hr)
head(hrv, 2)
summary(hrv)
head(p, 10)
summary(p)
head(scale, 5)
head(sleep, 2)
summary(sleep)
head(survey, 2)
summary(survey)
head(wear, 2)
summary(wear)
head(weather, 2)
summary(weather)
```

Average across times for each user and choose quantitative variables
```{r}
# I have chosen to use only diastolic and systolic for the preliminary run, because the other categories have a lot of missing data
bp_mean <- aggregate(cbind(diastolic,systolic) ~ user_code, data = bp, FUN = mean, na.rm = TRUE)
head(bp_mean)

#Removed is_resting data due to categorical nature of the variable
hr_mean <- aggregate(cbind(heart_rate) ~ user_code, data = hr, FUN = mean, na.rm = TRUE)
head(hr_mean)

#I have chosen to leave out feelings, mood, and sleep, due to probability of high missingness
hrv_mean <- aggregate(cbind(bpm, meanrr, mxdmn, sdnn, rmssd, pnn50, mode, amo, lf, hf, vlf, lfhf, total_power) ~ user_code, data = hrv, FUN = mean, na.rm = TRUE)
head(hrv_mean)

#Removed gender, city, country, because these are not quantitative variables. Also removed symptoms onset for now since we are not visualizing time data.
p$age_range <- as.factor(p$age_range)
str(p$age_range)
p_mean <- aggregate(cbind(height, weight, age_range) ~ user_code, data = p, FUN = mean, na.rm = TRUE)
head(p_mean,10)

#Did not include information from scales, because this is just a description of the data, not measured observations.

#Used only sleep duration because of the missingness in the other variables
sleep_mean <- aggregate(sleep_duration ~ user_code, data = sleep, FUN = mean, na.rm = TRUE)
head(sleep_mean)

#Used only value, because it was the only quantitative variable and served as an indication of covid symptom severity in general
survey_mean <- aggregate(value ~ user_code, data = survey, FUN = mean, na.rm = TRUE)
head(survey_mean)

#I did not include wearables data, as the amount of missingness indicated to me that most participants either did not have access to wearable technology or did not wear them often.

weather_mean <- aggregate(cbind(avg_temperature_C, atmospheric_pressure, precip_intensity, humidity, clouds) ~ user_code, data = weather, FUN = mean, na.rm = TRUE)
head(weather_mean)


```
Aggregate the datasets by user
```{r}

#Discluded sleep mean, because this significantly reduced the number of participants 

ag <- merge(bp_mean, hr_mean, by = "user_code")
ag <- merge(ag, hrv_mean, by = "user_code")
ag <- merge(ag, p_mean, by = "user_code")
ag <- merge(ag, survey_mean, by = "user_code")
ag <- merge(ag, weather_mean, by = "user_code")
ag

write.csv(ag, "/Users/sophialeeman/Documents/BME 580/BME-580/Aggregate_Data.csv", row.names=FALSE)
```
Scaling the data and computing PCA
```{r}
ag_scale <- ag[ -c(1) ]
pca <- prcomp(ag_scale, center = TRUE, scale=TRUE)
pca
```
Let's see the results!
```{r}
graph <- fviz_eig(pca, addlabels = T) 
graph
```
I would chose to use 4 dimensions by analysis of the location of the elbow in the graph.
```{r}
varPercent <- pca$sdev^2/sum(pca$sdev^2)*100
cumVar <- sum(varPercent[1:4])
cumVar
```
This accounts for 77% of the variance.

Display the most important features
```{r}
loadings <- pca$rotation
loadings <- as.data.frame(loadings[,1:4])
loadings  %>% arrange(desc(.372*abs(PC1)+.177*abs(PC2)+.153*abs(PC3)+.07*abs(PC4)))

```

```{r}
fviz_pca_ind(pca,
             col.ind = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     
             )
```

Visualize dimensions
```{r}
fviz_pca_biplot(pca)
```
```{r}
fviz_pca_var(pca,
             col.var = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE   
             )
```

```{r}
library(PCAtest)
result<-PCAtest(ag_scale, 100, 100, 0.05, varcorr=FALSE, counter=FALSE, plot=TRUE)

#According to PCAtest:
#"""The first 3 PC axes are significant and account for 70.2% of the total variation

#Variables 2, 6, 7, 8, 9, 11, 12, 13, 14, 16, 17, and 21 have significant loadings on PC 1
#Variables 2, 10, 18, 19, 24, and 25 have significant loadings on PC 2
#Variables 3, 4, 5, 10, 20, and 23 have significant loadings on PC 3"""

#https://pubmed.ncbi.nlm.nih.gov/35194531/, https://arleyc.github.io/PCAtest/ 

#This would indicate that systolic, mxdmn, sdnn, rmssd, pnn50, amo, lf, hf, vlf, total power, height, and average temperature are most important for the first dimension. Systolic, mode, weight, age range, humidity, and clouds are most important for the second dimension. And heart rate, bpm, meanrr, mode, value, and precipitation intensity are most significant for PC 3. This seems to line up with the PCA by prcomp. 
```